Verifying Cache Coherence Protocols for an infinite family of cache hierarchies

### 2014-02-06 paper/Abstract.tex
this the the correct version

Correct implementation of cache coherent systems is of high importance in
industrial use. We first provide a concise mathematical definition of a cache
coherent system as one which obeys store atomicity. We give a Coq-based proof
of an MSI cache coherence protocol for an infinite set of cache hierarchies. We
specify a set of invariants which imply store atomicity, enabling other
protocols to be verified for store atomicity automatically as long as they
satisfy the invariants. Our key insight for proving these invariants is a set
of fundamental properties.  We believe that these fundamental properties hold
for many other invalidation-based protocols with no silent drops -- we show
that informally for MESI, MOSI and MOESI protocols.

>>>>

%Correct implementation of cache coherent systems is of high importance in
%industrial use.
We provide a concise mathematical definition of a cache coherent system as one
which obeys store atomicity. We give a Coq-based proof of an MSI cache
coherence protocol for an infinite family of cache hierarchies. We specify a
set of invariants which imply store atomicity, enabling other protocols to be
verified for store atomicity automatically as long as they satisfy the
invariants. Our key insight for proving these invariants is a set of
fundamental properties.  We believe that these fundamental properties hold for
many other invalidation-based protocols with no silent drops -- we show that
informally for MESI, MOSI and MOESI protocols.

### 2014-02-06 paper/Background.tex
this the the correct version

\section{Correctness of a Cache Coherence Protocol}
\label{Sec:Background}

@@

Intuitively, the simplest view of shared memory is as a memory (array of addressible objects)
multiplexed amongst the many individual clients/processors, which send read/write
messages to the memory as needed. In
this view, all operations in the system appear to have a single unique total
ordering.

All modern systems have caches between the processor and memory for faster
memory access.  It is often also useful to permit some reorderings of requests
from a processor to the memory to improve performance.  For instance, in the
context of a processor with local store buffer, we may wish to allow a store to
bypass a subsequent load request to a different location. The specification of
which reorderings are permitted or restricted defines the memory consistency
model.

Most memory consistency models can be partitioned into two
parts~\cite{Arvind-memory-model}: (a) a ``global'' store-atomicity property
stating that for each location/address in memory, every read and write access
to that location occurs in some total order, with each operation seeing the
effects of previous operations and not later ones (in other words, each
location behaves like an atomic register) and (b), a ``local'' reordering
restriction between requests to different locations within each processor.
Changing the memory model by modifying the reordering restrictions will not
incur a reverification of the store-atomicity property.

@@

\subsection{Formal Definition of Store Atomicity}
We will model an abstract memory subsystem, which interfaces with a set of
processors. The requests issued by the processors to the memory subsystem can
be modeled using a function \reqFn, where $\reqFn(p, i)$ is the $i^{th}$ request
issued by processor $p$.

%Figure~\ref{req-resp} shows the
%data-type corresponding to a request issued from the processor to the
%memory subsystem.

We define a record type modeling a request in terms of a set of named fields.
A request contains the address (\addrQ) corresponding to the request, as well as a
\desc{} indicating whether the request is a load or a store. A store request
also contains the data supplied by the processor for that address (\dataQ). A
response (modeled with a different record type) contains a label (\labelR)
to identify the request uniquely, via a pair of a processor name and a processor-local
sequence number.  In the case of a load request, the memory
subsystem returns \dataR, the data returned by the memory subsystem for that
address. The memory subsystem also returns $\timeR \in \mathbb{N}$, which
intuitively represents the time at which the request is handled among the
global set of handled requests from all the processors. The \timeR{} field is
present solely as a specificational device to capture store atomicity, and it
would not be present in a running system implementation.

%We model request of a processor formally with the function
%\reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
%processor $p$, \ie{} $\reqFn: (\textit{Processor} * \mathbb{N}) @->
%\Request$.  The processor name and the position of the request within
%the processor uniquely identifies a request.

%\begin{figure}
%\centering
%\begin{subfigure}{5.5cm}
%\begin{tabular}{|lp{4.5cm}|}
%\hline
%\multicolumn{2}{|c|}{\Request}\\
%\hline
%\addrQ:& Address associated with the request\\
%\desc:& Request type (\Ld{} or \St)\\
%\dataQ:& Data associated with a \St{} request\\
%\hline
%\end{tabular}
%\end{subfigure}~~~~
%\begin{subfigure}{5.5cm}
%\begin{tabular}{|lp{4.5cm}|}
%\hline
%\multicolumn{2}{|c|}{\Response}\\
%\hline
%\labelR:& Label identifying the request corresponding to this response\\
%\timeR:& Natural number representing global order of (processing) requests\\
%\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
%\hline
%\end{tabular}
%\end{subfigure}
%\caption{Data types for request and response}
%\label{req-resp}
%\end{figure}

We model a single protocol execution as a set of all the requests and responses
that sent in the course of that execution.
The fundamental parameter is a (potentially infinite) set \Response{} of all responses
that the memory sends to processors.  From here, the associated requests follow by
applying the function $\reqFn$ to map each response to the request that provoked it.
Responses should obey the following sanity conditions. 
\begin{defn} (No label is reused across multiple distinct responses.)
%Since \dataR{} field is
%irrelevant for store requests, so uniqueness of responses does not depend on
%\dataR{} field for stores.)
\small
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),\\
r_1.\labelR = r_2.\labelR @-> r_1.\timeR = r_2.\timeR \wedge\\
(\reqFn(r_1.\labelR) = \Ld @-> r_1.\dataR = r_2.\dataR)
\end{multline*}
\label{uniqRespLabels}
\end{defn}

\begin{defn} (No response can share a \timeR{} value with another store response.)
\small
\begin{multline*}
\textit{uniqStTimes} := 
\forall (r_1, r_2 \in \Response),
r_1.\timeR = r_2.\timeR @-> \\
\reqFn (r_1.\labelR).\addrQ = \reqFn (r_2.\labelR).\addrQ @->\\
\reqFn (r_1.\labelR).\desc = \St @->
r_1.\labelR = r_2.\labelR
\end{multline*}
\label{uniqStTimes}
\end{defn}

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.

\begin{defn}
\small
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response), r_1.\labelR.\fst = r_2.\labelR.\fst @->\\
r_1.\labelR.\snd < r2.\labelR.\snd @-> r_1.\timeR \le r_2.\timeR
\end{multline*}
\label{localOrdering}
\end{defn}

This brings us to the main condition for store atomicity, which states that the
data returned for a load response corresponds to the \emph{last earlier} store
request from any processor that has been handled for that address,
\emph{last} and \emph{earlier} being defined according to the field \timeR{}
returned in the responses.  This proposition is indicated by
\textit{lastEarlierValue}$(a, t, d)$ for a load request for address $a$ handled
at time $t$, returning data $d$.  In case no store has happened earlier for a
particular address when a load is processed, then the load response returns the
initial value for the address, the whole proposition indicated by
\textit{initialValue}$(a, t, d)$.
\begin{defn}
\small
\begin{multline*}
\textit{initialValue}(a, t, d) := d = \initData (a) \; \wedge \\
\forall (r' \in \Response), 0 \le r'.\timeR < t @->\\
\neg (\reqFn (r'.\labelR).\addrQ = a \wedge \reqFn (r'.\labelR).\desc = \St)
\end{multline*}
\label{initialValue}
\end{defn}

\begin{defn}
\small
\begin{multline*}
\textit{lastEarlierValue}(a, t, d) := \exists (r_m \in \Response),\\ \reqFn (r_m.\labelR).\dataQ = d \wedge 
\reqFn (r_m.\labelR).\desc = \St \; \wedge\\ \reqFn (r_m.\labelR).\addrQ = a \wedge
\forall (r' \in \Response), r_m.\timeR < r'.\timeR < t @->\\
\neg (\reqFn (r'.\labelR).\addrQ = a \wedge \reqFn (r'.\labelR).\desc = \St)
\end{multline*}
\label{lastEarlierValue}
\end{defn}

Using the above definitions, we can define the store atomicity theorem as follows:
\begin{thm}
\small
\begin{multline*}
\textit{storeAtomicity} := 
\textit{uniqRespLabels} \wedge
\textit{uniqStTimes} \wedge \textit{localOrdering} \; \wedge \\
(\forall (r \in \Response), \mylet q := \reqFn(r.\labelR) \myin q.\desc = \Ld \; @-> \\
(\textit{initialValue}(q.\addrQ, r.\timeR, r.\dataR) \vee \textit{lastEarlierValue}(q.\addrQ, r.\timeR, r.\dataR)))
\end{multline*}
\label{storeAtomicity}
\end{thm}

>>>>

\section{Correctness of a Cache Coherence Protocol}
\label{Background}

@@

Intuitively, the simplest view of shared memory is as an array of addressible
objects in which a single entry can be read or written by one of the many
processors at a time.  In this view, all operations in the system appear to
have a single unique total ordering.

All modern systems have caches between the processor and memory for faster
memory access. It is often also useful to reorder requests from a processor to
the memory to improve performance. For instance, a load may bypass a previous
store to a different location. The \emph{memory consistency model} describes
the semantics of a system in the presence of caches and reorderings.

Most memory consistency models can be partitioned into two
parts~\cite{Arvind-memory-model}: (a) a ``global'' store-atomicity property
stating that for each location/address in memory, every read and write access
to that location from any processor, occurs in some total order, with
each operation seeing the effects of previous operations and not later ones (in
other words, each location behaves like an atomic register) and (b), a
``local'' reordering restriction between requests to different locations within
each processor. Changing the memory model by modifying the reordering
restrictions will not incur a reverification of the store-atomicity property.

@@

We will now provide the formal definition of store atomicity.

\subsection{Formal Definition of Store Atomicity}
A memory subsystem can be modeled as the following function. Its input models a
set of infinite number of requests ever sent by any processor, and its output
models a set of (possibly infinite number of) responses ever sent by the memory
subsystem to any processor.

Each input to the memory subsystem is isomorphic to a function
$\reqFn:(Processor*\mathbb{N}) @-> \Request$, where $\reqFn(p, i)$ models the
$i^{th}$ request issued by processor $p$. An element of the range-set of
\reqFn{}, \viz \Request{} contains fields (a) address (\addrQ) corresponding to
the request, (b) \desc{} indicating whether the request is a load or a store
and (c) data (\dataQ) containing the data supplied by the processor for stores.

Each output of the memory subsystem contains a set (called \Response) of
elements containing the fields (a) \labelR, containing a pair from the set
$Processor*\mathbb{N}$, which is meant to identify a request uniquely, (b)
\dataR, meant to model the data returned from the memory subsystem and (c)
\timeR, a value from set $\mathbb{N}$, meant to represent the ``time'' at which
the corresponding request (identified by \labelR) is handled by the memory
subsystem. \timeR{} does not have a counterpart in a real implementation, it is
present solely as a specificational device to capture store atomicity. Since a
request corresponding to a response is identified using the field \labelR, for
notational convenience, we will access the fields of the corresponding request
directly from the response. For instance, $r.\addrQ$ for $r \in \Response$
denotes the $(r.\labelR).\addrQ$, \etc.

If \reqFn{} is an input to the memory subsystem, and the set \Response{} is the
produced output, then the memory subsystem is said to be store atomic if it
obeys Theorem \fullref{storeAtomicity} defined as follows. We use the word
response to denote an element in the set \Response{}.
%
\begin{defn} (Responses have unique labels)
\small
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),
r_1.\labelR = r_2.\labelR @-> \\ r_1.\timeR = r_2.\timeR \wedge
r_1.\dataR = r_2.\dataR
\end{multline*}
\label{uniqRespLabels}
\end{defn}
%
\begin{defn} (No response can share a \timeR{} value with another store
response for the same address.)
\small
\begin{multline*}
\textit{uniqStTimes} := 
\forall (r_1, r_2 \in \Response),
r_1.\timeR = r_2.\timeR @-> \\
r_1.\addrQ = r_2.\addrQ @->
r_1.\desc = \St @->
r_1.\labelR = r_2.\labelR
\end{multline*}
\label{uniqStTimes}
\end{defn}
%
\begin{defn} (The global ordering of responses to the same processor should not
violate the ordering of the corresponding requests within the processor)
\small
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response), r_1.\labelR.\fst = r_2.\labelR.\fst @->\\
r_1.\labelR.\snd < r_2.\labelR.\snd @-> r_1.\timeR \le r_2.\timeR
\end{multline*}
\label{localOrdering}
\end{defn}
%
\begin{defn} (If a response to a particular request from a processor exists in the \Response{} set,
then responses for all previous requests from that processor must exist)
\small
\begin{multline*}
\textit{allPrevious} :=
\forall (r \in \Response), \forall i < r.\labelR.\snd, \\
\exists (r' \in \Response) \text{ such that } r'.\labelR.\fst = r.\labelR.\fst \wedge r'.\labelR.\snd = i
\end{multline*}
\label{allPrevious}
\end{defn}
%
This brings us to the main condition for store atomicity, which states that the
data returned for a load response corresponds to the \emph{last earlier} store
request from any processor that has been handled for that address, \emph{last}
and \emph{earlier} being defined according to the field \timeR{} returned in
the responses.  This proposition is indicated by \textit{lastEarlierValue}$(r)$
for a load request $r$.  In case no store has happened earlier for a particular
address when a load is processed, then the load response returns the initial
value for the address ($\initData(a)$), the whole proposition indicated by
\textit{useInitialValue}$(r)$.
\begin{defn}
\small
\begin{multline*}
\textit{useInitialValue}(r) := r.\dataR = \initData(a) \; \wedge \neg \exists (r' \in \Response), \\
r'.\timeR < t \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
\end{multline*}
\label{useInitialValue}
\end{defn}
%
\begin{defn}
\small
\begin{multline*}
\textit{lastEarlierValue}(r) := \exists (r_m \in \Response), r_m.\dataQ = r.\dataQ \; \wedge 
\\ r_m.\desc = \St \wedge r_m.\addrQ = r.\addrQ \wedge
\neg \exists (r' \in \Response), \\
r_m.\timeR < r'.\timeR < t \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
\end{multline*}
\label{lastEarlierValue}
\end{defn}
%
Using the above definitions, we can define the store-atomicity theorem as follows:
\begin{thm}\textit{storeAtomicity}:
\small
\begin{multline*}
\textit{uniqRespLabels} \wedge
\textit{uniqStTimes} \wedge \textit{localOrdering} \wedge \textit{allPrevious} \; \wedge \\
(\forall (r \in \Response), r.\desc = \Ld \; @-> \\
(\textit{useInitialValue}(r) \vee \textit{lastEarlierValue}(r)))
\end{multline*}
\label{storeAtomicity}
\end{thm}

### 2014-02-06 paper/System.tex
this the the correct version

Input and output happens at the leaves in the form of memory requests
issued from and responses issued to the associated processor. These will be
represented by two mappings, one from $\mathbb{N}$ to requests and the other
from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
position of the next request to process; it is set to $0$ initially.
\adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
\adamc{I hope the diagram explains what these mappings mean, and in general what
the tree structure really means.}

>>>>

Each node contains the fields mentioned in Figure \ref{hier}. We will
explain what the fields mean in the following discussion.

\begin{figure}
\includegraphics[scale=.5]{hier}
\caption{A cache hierarchy, and the fields in each node}
\label{hier}
\end{figure}

Input and output happens at the leaves in the form of memory requests
issued from and responses issued to the associated processor. These will be
represented by two mappings, one from $\mathbb{N}$ to requests and the other
from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
sequence number of the next request to process; it is set to $0$ initially.
\adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
\adamc{I hope the diagram explains what these mappings mean, and in general what
the tree structure really means.}
